---
title: "Analysis of Weather Events impacting Health and Economy in USA"
author: "Yang Yuzhong"
date: "3/18/2015"
output: html_document
---

## Synopsis
This report outlines the analysis done to determine the impact of various 
weather events on the economy and health of United States. The data used in 
this report is based on the 1950 - 2011 storm database collected from 
U.S. National Oceanic and Atmospheric Administration’s (NOAA). 

Due to fewer events recorded for years earlier than 1996, the analysis
was only performed on years between 1996-2011. The metrics for determining 
weather event impacts are `Fatalities` and `Injuries` for health impact and 
`Property Damage` and `Crop Damage` for economic impact.

The analysis results has shown that *Tornado* caused the highest number of 
casualties and *Flood* caused the highest economic damage to United States 
between 1996-2011. Also, *Flood* related events are highly ranked in terms 
of impact to United States' health and economy.

## Data Processing
### Setting Up of Dataset and Environment
```{r calllibraries}
library(R.utils)
library(ggplot2)
library(dplyr)
```

```{r downloadfile}
setwd("~/workspace//Reproducible Research/RepData_PeerAssessment2/")

if (!"repdata_data_StormData.csv.bz2" %in% dir()) {
    download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
        destfile = "repdata_data_StormData.csv.bz2")
    bunzip2("repdata_data_StormData.csv.bz2", overwrite=T, remove=F)
}

if ("repdata_data_StormData.csv" %in% dir()) {
    data<-read.csv("repdata_data_StormData.csv")
    names(data)<-tolower(names(data))
}
```

### Filtering the Dataset
The dataset is read in and all headers converted to lowercase with `tolower()`
and `names()`. Next, `strptime()` and `format()` is performed on _bgn_date_ 
to extract the year for each entry in the dataset. _year_ column is then added
to the original dataset.

```{r formatyear}
datecol<-strptime(data$bgn_date, "%m/%d/%Y %H:%M:%S", tz="UTC")
datecol<-format(datecol, "%Y")

data<-mutate(data,year=datecol)
```

Following this, _datayear_ was created to hold the count of records for each 
year in the dataset.Based on information from 
[Details of Storm Events](http://www.ncdc.noaa.gov/stormevents/details.jsp),
years earlier than 1996 were filtered from the dataset with `%in%` operator so 
that we perform analysis on years that have the most holistic types of 
weather events.

```{r filteryear}
datayear<-data %>%
    group_by(year) %>%
    summarize(entries = length(evtype))

datayear<-datayear[datayear$year>1995,]

data<-filter(data,data$year %in% datayear$year)
```

Lastly, events without any fatalities, injuries, crop and property dmg are
removed as they are not required to answer the questions for the analysis.

```{r filternoimpact}
data<-filter(data,fatalities != 0 | injuries != 0 | propdmg != 0 | cropdmg != 0)
```

### Determining health impact for each record
To determine health impact by event type, a new column _totaldamage_ that is 
the sum of fatalities and injuries from event is added to the dataset using 
`mutate()`.

```{r sumhealth}
data<-data %>%
    mutate(totalimpact=(data$fatalities + data$injuries))
```

### Determining economic impact for each record
To determine economic impact by event type, the true value of crop damage,
derived from _cropdmg_ and _cropdmgexp_ containing the exponential multplier. 
This is the same for property damage with _propdmg_ and _propdmgexp_. Then the
total economic impact is obtained by the sum of crop and property damage.

To do this, codebook from
[Investigative Reporters & Editors - Record Layout](http://ire.org/media/uploads/files/datalibrary/samplefiles/Storm%20Events/layout08.doc) were used for the character symbols in 
_cropdmgexp_ and _propdmgexp_, were used to convert them into numeric format.

```{r subdmg}
data$cropdmgexp<-tolower(as.character(data$cropdmgexp))
data$cropdmgexp[data$cropdmgexp == "b"]<-"9"
data$cropdmgexp[data$cropdmgexp == "m"]<-"6"
data$cropdmgexp[data$cropdmgexp == "k"]<-"3"
data$cropdmgexp[data$cropdmgexp == "h"]<-"2"
data$cropdmgexp[data$cropdmgexp == ""]<-"0"
data$cropdmgexp[is.na(data$cropdmgexp)]<-"0"
data$cropdmgexp<-as.numeric(data$cropdmgexp)

data$propdmgexp<-tolower(as.character(data$propdmgexp))
data$propdmgexp[data$propdmgexp == "b"]<-"9"
data$propdmgexp[data$propdmgexp == "m"]<-"6"
data$propdmgexp[data$propdmgexp == "k"]<-"3"
data$propdmgexp[data$propdmgexp == "h"]<-"2"
data$propdmgexp[data$propdmgexp == ""]<-"0"
data$propdmgexp[is.na(data$propdmgexp)]<-"0"
data$propdmgexp<-as.numeric(data$propdmgexp)
```

After the conversion, a new column _totaldamage_ that is the sum of crop and
property damage is added to the dataset using `mutate()`.

```{r sumdmg}
data<-data %>%
    mutate(totaldamage=(data$cropdmg * 10^data$cropdmgexp) + 
               (data$propdmg * 10^data$propdmgexp))
```

### Merging Similar Event types
As the event types given in the dataset are too inconsistently named and manual 
merging of the event types into one of the 48 event types reflected in
[Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
would be too tedious, a partial merging of similar event types that have high
health and economic impact is performed.

Firstly, to determine the event types to merge, _healthimpact_ and 
_economicimpact_ are created by summing up _totalimpact_ and _totaldamage_
columns respectively then arranged in descending order. This helps in
determining the top 20 high health and economic impact event types to perform
merging to reduce the scope of merging similar event types.

```{r topimpacts}
tophealth<- data %>%
    group_by(evtype) %>%
    summarize(totalimpact=sum(totalimpact)) %>%
    arrange(desc(totalimpact))

head(tophealth,n=20)

topeconomic<-data %>%
    group_by(evtype) %>%
    summarize(totaldamage=sum(totaldamage)) %>%
    arrange(desc(totaldamage))

head(topeconomic,n=20)
```

With the information on the top 20 health and economic impact event types,
similar event types were merged under one label using `grep()`.

```{r evtypeclassification}
data$evtype<-as.character(data$evtype)
data$evtype[grep("non.tstm",data$evtype,ignore.case=TRUE)]<-"STRONG WIND"
data$evtype[grep("thunderstorm|tstm",data$evtype,ignore.case=TRUE)]<-"THUNDERSTORM"
data$evtype[grep("wind",data$evtype,ignore.case=TRUE)]<-"STRONG WIND"
data$evtype[grep("^strong.wind|gusty.wind|^high.wind|^wind",
                 data$evtype,ignore.case=TRUE)]<-"STRONG WIND"
data$evtype[grep("flash flood",data$evtype,ignore.case=TRUE)]<-"FLASH FLOOD"
data$evtype[grep("^coastal flood|tidal flood|flooding/erosion",
                 data$evtype,ignore.case=TRUE)]<-"COASTAL FLOOD"
data$evtype[grep("flood",data$evtype,ignore.case=TRUE)]<-"FLOOD"
data$evtype[grep("rip",data$evtype,ignore.case=TRUE)]<-"RIP CURRENT"
data$evtype[grep("heat",data$evtype,ignore.case=TRUE)]<-"HEAT"
data$evtype[grep("fire",data$evtype,ignore.case=TRUE)]<-"WILDFIRE"
data$evtype[grep("freeze|cold",data$evtype,ignore.case=TRUE)]<-"COLD"
data$evtype[grep("surge",data$evtype,ignore.case=TRUE)]<-"STORM SURGE"
data$evtype[grep("hail",data$evtype,ignore.case=TRUE)]<-"HAIL"
```

After the merging of similar event types, _healthimpact_ and _economicimpact_
is created by summing of _totalimpact_ and _totaldamage_ columns respectively
then arranged in descending order.

```{r finalimpacts}
healthimpact<- data %>%
    group_by(evtype) %>%
    summarize(totalimpact=sum(totalimpact)) %>%
    arrange(desc(totalimpact))

economicimpact<-data %>%
    group_by(evtype) %>%
    summarize(totaldamage=sum(totaldamage)/10^6) %>%
    arrange(desc(totaldamage))
```

## Results
After preprocessing the data, analysis are performed to determine weather 
events types that are most harmful w.r.t Population Health and had the
greatest economic impact.

### Events Most Harmful with Respect to Population Health
```{r healthplot}
healthdata<-healthimpact[1:10,]
healthdata$evtype<-factor(healthdata$evtype,
                          level = as.character(healthdata$evtype))
healthplot <- ggplot(data = healthdata, aes(x = evtype, y = totalimpact)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
    geom_bar(stat="identity") +
    ggtitle("Most Harmful Weather Event Types in USA (1996-2011)") +
    xlab("Weather Event Type") +
    ylab("No. of Casualties")
healthplot
```

From the health impact barplot, *Tornado* caused the highest number of 
casualties between 1996-2011. This is almost double that of *Flood* and *Heat* 
related weather events.

### Events with Greatest Economic Consequences
```{r econplot}
econdata<-economicimpact[1:10,]
econdata$evtype<-factor(econdata$evtype, 
                        level = as.character(econdata$evtype))
econplot<- ggplot(data = econdata, aes(x = evtype, y = totaldamage)) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
    geom_bar(stat="identity") +
    ggtitle("Costliest Weather Event Types in USA (1996-2011)") +
    xlab("Weather Event Type") +
    ylab("Property/Crop Damage (in $USD Millions)")
econplot
```

From the economic impact barplot, *Flood* caused the largest Property and Crop
related damage. This is almost triple that of *Hurricane* and *Storm Surge* 
related weather events.

### Conclusion
*Flood* related weather events rank highly in terms of economic impact and 
number of casualty. *Tornado* is 4th in terms of property/crop damage although 
it causes the highest number of casualties.
